### 2010-07-22 papers/icfp-2010-desc/paper_idesc.tex
gaol: boiling the indexed examples

@@

Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrized
by \(\V{I}\), but indexed trivially.
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \SYMBDSigma \\
\quad
 \EnumT\bigRedBracket{\begin{array}{r}
                                          \SYMBDVar\\
                                          \SYMBDConst\\
                                          \DProd{}{}\\
                                          \SYMBDSigma\\
                                          \SYMBDPi
                       \end{array}}
         \; \bigRedBracket{\begin{array}{l@{}l}
  (\DProd{\DConst{\V{I}} &}{\DConst{\Unit}})                  \\
  (\DProd{\DConst{\Set}  &}{\DConst{\Unit}})                  \\
  (\DProd{\DVar{\Void}}{\DProd{\DVar{\Void}&}{\DConst{\Unit}}})  \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
                                   \end{array}}
}\]

@@

One way to code constrained datatypes is to appeal to a suitable
notion of propositional equality \(\PropEq\) on indices. The
constraints are expressed as `Henry Ford' equations in the datatype.
For vectors:
%
\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\VecD\: \V{X}\: \V{i} = \SYMBDSigma\\
\quad
\EnumT{\bigRedBracket{\begin{array}{@{}r@{}}
\VecNil\\ \SYMBVecCons
\end{array}}}
\; \bigRedBracket{
                          \begin{array}{@{}r@{}}
                            (\DConst{(\NatZero\PropEq\V{i})}) \\
 ( \DSigma{\Nat}{\LAM{\V{n}}
   \DProd{\DConst{\V{X}}}
     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                          \end{array}
                          }
}\]

@@

We have been careful to keep our setup agnostic with respect to
notions of propositional equality. Any will do, according to your
convictions, or for vectors, none---equality for \(\Nat\) is definable by
recursion---and many variations are popular. The traditional
homogeneous identity type used in Coq is not adequate to support
dependent pattern matching, but its heterogeneous variant, allowing
equations between elements of arbitrary types, is sufficient to
allow the translation of structurally recursive pattern matching
programs to \(\F{indI}\)~\cite{goguen:pattern-matching}. Our present
inclination is towards the extensional equality proposed by
\citet{altenkirch:ott}, which also sustains the translation.

@@

However, sometimes, we can actually remove these equations altogether. Let us
look back at $\D{Vec}$. We note that the equations are introduced
because we are \emph{storing} the index of the inductive family.
However, \emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}.  By examining the
incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
optimisations to our initial definition of $\D{Vec}$. This gives the
following, equivalent definition:
%
\[\stk{
\VecD \:\PITEL{\V{X}}{\Set} : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\VecD\:\V{X}\: \NatZero     & \mapsto & \DConst{\Unit} \\
\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto &
 \DProd{\DProd{\DConst{\V{X}}}{\DVar{\V{n}}}}{\DConst{\Unit}}
\end{array}

}\]
%
The equations (and constructors) have simply disappeared. A similar
example is $\SYMBFin$, specified by:
%
\[
\stk{
\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}\]
%
In this case, we can apply forcing, but not detagging, since both
$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
%
\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll@{}l}
\FinD\: \NatZero         & \mapsto & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & \mapsto & \SYMBDSigma & (\EnumT{\sqr{ \FinZero \: \SYMBFinSuc }}) \\
                         &         &             & \bigRedBracket{
                                                   \begin{array}{l}
                                                       \DConst{\Unit} \\
                                                       \DVar{\V{n}}
                                                   \end{array}}
\end{array}
}\]

Note that forcing a description is not guaranteed to remove all
constraints. It is subject to future work to see if constraints can be
entirely eradicated, or presented more conveniently to the developer.
Finally, it is worth mentioning that these optimisations are
\emph{source-to-source} transformations on descriptions.

@@

When defining an indexed datatype, we have access to its
index. Therefore, we can use this index to influence the choice of
constructors. This captures the essence of dependent datatypes: a
term -- the index -- has the ability to influence the datatype. We
define tagged indexed descriptions to capture this specificity.

We divide a tagged indexed description in two parts: first, the
constructors that do not depend on the index; then, the constructors
that do. The non-dependent part mirrors the definition for non-indexed
descriptions. The index-dependent part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to enable or disable constructors.
%
\[
\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
 \ITagIDesc{\V{I}} & 
     \SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
\end{array}
\]

@@

We use the notation $\toIDesc{\V{D}}$ to denote the indexed
description computed from the tagged indexed description $\V{D}$. Its
expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
descriptions. For brevity, we shall also overload $\IMu{ty}{D}$ for
$\IMu{ty}{(\toIDesc{D})}$, as the bidirectional presentation let us
resolve the ambiguity.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   {\sqr{\SYMBEval\: \SYMBEcond \,}} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \sqr{\sqr{\SYMBEplus} \quad \sqr{\SYMBEle}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

>>>>

\newcommand{\vtup}[2]{\bigRedBracket{\begin{array}{@{}#1@{}}#2\end{array}}}

@@

Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrized
by \(\V{I}\), but indexed trivially.
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \SYMBDSigma \\
\quad
 \EnumT\vtup{r}{\SYMBDVar\\
                \SYMBDConst\\
                \DProd{}{}\\
                \SYMBDSigma\\
                \SYMBDPi}
         \; \vtup{l@{}l}{
  (\DProd{\DConst{\V{I}} &}{\DConst{\Unit}})                  \\
  (\DProd{\DConst{\Set}  &}{\DConst{\Unit}})                  \\
  (\DProd{\DVar{\Void}}{\DProd{\DVar{\Void}&}{\DConst{\Unit}}})  \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
                                   }
}\]

@@

One way to code constrained datatypes is to appeal to a suitable
notion of propositional equality \(\PropEq\) on indices. The
constraints are expressed as `Henry Ford' equations in the datatype.
For vectors:
%
\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\VecD\: \V{X}\: \V{i} = \SYMBDSigma\\
\quad
\EnumT{\vtup{r}{\VecNil\\ \SYMBVecCons}}
\; \vtup{r}{
                            (\DConst{(\NatZero\PropEq\V{i})}) \\
 ( \DSigma{\Nat}{\LAM{\V{n}}
   \DProd{\DConst{\V{X}}}
     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                          }
}\]

@@

In this paper, we remain anxiously agnostic about
propositional equality. Any will do, according to
conviction; many variations are popular. The
homogeneous identity type used in Coq is ill-suited to
dependent types, but its heterogeneous variant (forming equations
regardless of type) allows the translation of pattern
matching with structural recursion to
\(\F{indI}\)~\cite{goguen:pattern-matching}. The
extensional equality of \citet{altenkirch:ott} also sustains the translation.

@@

However, sometimes, the equations are redundant. 
Looking back at $\D{Vec}$, we find that the equations constrain
the choice of constructor and stored tail index retrospectively.
But \emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}!  If we
analyse the incoming index, we can tidy our description of $\D{Vec}$
as follows:
%
\[\stk{
\VecD \:\PITEL{\V{X}}{\Set} : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\VecD\:\V{X}\: \NatZero     & \mapsto & \DConst{\Unit} \\
\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto &
 \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
\end{array}

}\]
%
The constructors and equations have simply disappeared. A similar
example is $\SYMBFin$ (bounded numbers), specified by:
%
\[
\stk{
\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}\]
%
In this case, we can eliminate equations but not constructors, since both
$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
%
\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma
 {\EnumT{\vtup{r}{\FinZero \\ \SYMBFinSuc }}}
 {\vtup{l}{ (\DConst{\Unit}) \\
            (\DVar{\V{n}})}}
\end{array}
}\]

This technique of extracting information by case analysis on indices
applies to descriptions exactly where Brady's `forcing' and
`detagging' optimisations apply in compilation. They eliminate just
those constructors, indices and constraints which are redundant even
in \emph{open} computation. In \emph{closed} computation, where proofs
can be trusted, all constraints are dropped.

@@

Let us reflect this index analysis technique.
We can divide a description of tagged indexed data in two: first, the
constructors that do not depend on the index; then, the constructors
that do. The non-dependent part mirrors the definition for non-indexed
descriptions. The index-dependent part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to vary the `menu' of constructors.
%
\[
\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} \PI{\V{i}}{\V{I}} \spi{\V{E}}{\LAM{\_} \IDesc{\V{I}}} \\
 \ITagIDesc{\V{I}} & 
     \SIGMA{\V{F}}{\V{I} \To \EnumU} \PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{\LAM{\_} \IDesc{\V{I}}} 
\end{array}
\]

@@

We use the notation $\toIDesc{\V{D}}$ to denote the indexed
description computed from the tagged indexed description $\V{D}$. Its
expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
descriptions. For brevity, we shall also overload $\IMu{I}{D}$ for
$\IMu{I}{(\toIDesc{D})}$, as the bidirectional presentation let us
resolve the ambiguity.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \vtup{l}{
                   {\sqr{\SYMBEval\: \SYMBEcond \,}} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \sqr{\sqr{\SYMBEplus} \quad \sqr{\SYMBEle}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]
